### 2017-07-22 chapters/introduction.tex
Add latest

One of the earlies examples of modelling applied to cardiac
physiology dates back to 1892 when Robert Wood approximated the
ventricle as a thinned-wall sherical shell.
% In this way, \emph{wall
% stress} could be analytically estimated via the famous \emph{Law of Laplace}
% \cite{woods1892few}. This law relates the wall stress $\sigma$ to the endocardial
% pressure $P$, the endocarial radius $r$ and the wall thickness $w$
% and can in its simplest for be expressed as
% \begin{align}
%   \sigma = \frac{P \times r}{2 w}.
% \end{align}
% Wall stress can be thought of as the load that the heart must work
% against, and it is therefore important to understand how different
% quantities affects the working heart. For instance when the heart
% ejects blood, the load that the heart ejects blood against is
% typically referred to as \emph{afterload}, but this is really the same
% as the wall stress in that particular state. 
% Even though it is obvious that the shape of a ventricle in most cases
% clearly deviates from a sphere, this forumla provides clinician with a tool
% that can be used to understand how diseases affect the heart. Take for
% example the case of ventricular dilation, when the inner radius of 
% the ventricle increases, resulting in increased wall stress and
% hence the heart has to work harder in order to eject blood to the body. In the other
% end we have the case of concentric hypertrophy which results in wall
% thickening and hence reduced the wall stress. In this case more muscle
% fibers are recruited and therefore the tension experienced by each
% individual unit will be less. 
Since 1892 more realistic and sophisticated models have been developed
in order to better describe the anatomical as well as the biophysical
nature of the ventricle. The first extension came in 1927 to
thick-walled spheres \cite{love2013treatise}. In the 1960 the spheres
were substitued by ellipsoids, with thin-walled models in 1963
\cite{sandler1963left}, and by thick-walled models in 1968
\cite{wong1968stress}. A thourough comparison of the different early
models  is presented in \cite{huisman1980comparison}. The first
reported usage of image-based geometries dates back to 1972
\cite{gould1972vivo} were X-ray was used to image the left ventricle
and the finite element method was used to estimate wall stress. The relatively
poor computational power, together with the low image quality really
challenged the biomedical engineers at that time. The following
phrase, taken form \cite{heethaar1977computer} illustrates this fact:  
\emph{``Programming for the solution of the large number of
  simultaneous linear equations generated by the finite element method
  on computers with capacity not exceeding 32K words of memory
  required special attention to very compact storage of the
  stiffness matrix and the retrieval of its coefficients''}.
A paradigm shift in the field of cardiac mechanical modeling came in
the late 1980s at the University of Auckland. Here the first  continuum
description of the heart were developed \cite{hunter1988analysis},
which is still in use today.

@@

Today, advanced imaging techniques allow us to aquire detailed
mechanical information about the heart of the individual. In addition,
advanced numerical algorithms combined with high performance computing
power enable us to get more and more detailed simulation of a patients
heart.

@@

The need for building adequate patient-specific models that captures
the geometrical information observed in the images as well as the
underlying biophysical processes, is recognized as one of the key
challenges in modern bioengnieering \cite{hunter2010vision}.
In the future, models could be used to as a diagnostic tool to extract
features from a patients heart that are otherwise impossible to extract from
images. Eventually, models could potentially predict the outcome of
different treatment strategies and guide clinicians in the decision making.

In other parts of physical sciences and engineering, such as the
space industry or the weather forecast, the role of the computational
science has been cruicial for its development. Without the increasing
computational power predicted by Moore's law
\cite{brock2006understanding} and better numerical methods, such as the
finite element method, scientist would not have been able to predict
complex physical phenomena such as a Tsunami, or being able to send
humans out in space.

Despite our ability to simulate complex physical systems such as
the weather, we still have a long way to go before we are able to
simulate a human organ, such as the heart. The multiscale nature of
biophysical processes has challenged bioengnieers for decades.
Even though we understand processes at each level, such as the flow of
ions in and out of the cell, or the electical transduction through the
heart, the question of how to couple these levels together is still open
\cite{noble2002modeling}.

Another question is how to extract data
observed in medical images and constrain models, in order to make
models specific to the individual. These questions has to be adressed
before we can take the next step into personalized simulations for
\emph{in silico} medicine, and use computer simulations to guide
treatment, and predict outcome.

@@

The mechanics of the heart can be described via equations of nonlinear
elasticity which is traditionally formulated using a continuum
mechanics approach. This approach has been extensively used in other
parts of engineering, and numerical solutions to these partial
differential equations (PDE) are solved using the finite element
method (FEM) with the open source software package FEniCS.

>>>>

One of the earliest examples of modelling applied to cardiac
physiology dates back to 1892 when Robert Wood approximated the
ventricle as a thinned-wall spherical shell.
% In this way, \emph{wall
% stress} could be analytically estimated via the famous \emph{Law of Laplace}
% \cite{woods1892few}. This law relates the wall stress $\sigma$ to the endocardial
% pressure $P$, the endocarial radius $r$ and the wall thickness $w$
% and can in its simplest for be expressed as
% \begin{align}
%   \sigma = \frac{P \times r}{2 w}.
% \end{align}
% Wall stress can be thought of as the load that the heart must work
% against, and it is therefore important to understand how different
% quantities affects the working heart. For instance when the heart
% ejects blood, the load that the heart ejects blood against is
% typically referred to as \emph{afterload}, but this is really the same
% as the wall stress in that particular state. 
% Even though it is obvious that the shape of a ventricle in most cases
% clearly deviates from a sphere, this forumla provides clinician with a tool
% that can be used to understand how diseases affect the heart. Take for
% example the case of ventricular dilation, when the inner radius of 
% the ventricle increases, resulting in increased wall stress and
% hence the heart has to work harder in order to eject blood to the body. In the other
% end we have the case of concentric hypertrophy which results in wall
% thickening and hence reduced the wall stress. In this case more muscle
% fibers are recruited and therefore the tension experienced by each
% individual unit will be less. 
Since 1892 more realistic and sophisticated models have been developed
in order to better describe the anatomical as well as the biophysical
nature of the ventricle. The first extension came in 1927 to
thick-walled spheres \cite{love2013treatise}. In the 1960s the spheres
were substituted by ellipsoids, with thin-walled models in 1963
\cite{sandler1963left}, and with thick-walled models in 1968
\cite{wong1968stress}. A thorough comparison of the different early
models  is presented in \cite{huisman1980comparison}. The first
reported usage of image-based geometries dates back to 1972
\cite{gould1972vivo}, were X-ray was used to image the left ventricle
and the finite element method was used to estimate wall stress. The relatively
poor computational power, together with the low image quality really
challenged the biomedical bioengineers at that time. The following
phrase, taken form \cite{heethaar1977computer} illustrates this fact:  
\emph{``Programming for the solution of the large number of
  simultaneous linear equations generated by the finite element method
  on computers with capacity not exceeding 32K words of memory
  required special attention to very compact storage of the
  stiffness matrix and the retrieval of its coefficients''}.
A paradigm shift in the field of cardiac mechanical modeling came in
the late 1980s at the University of Auckland. Here the first models
based on continuum description of the heart were developed
\cite{hunter1988analysis}, which is still in use today.

@@

Today, advanced imaging techniques allow us to acquire detailed
mechanical information about the heart of the individual. In addition,
advanced numerical algorithms combined with high performance computing
power enable us to get more and more detailed simulation of a patients
heart.

@@

The need for building adequate patient-specific models that captures
the geometrical information observed in the images as well as the
underlying biophysical processes, is recognized as one of the key
challenges in modern bioengineering \cite{hunter2010vision}.
In the future, models could be used to as a diagnostic tool to extract
features from a patients heart that are otherwise impossible to extract from
images. Eventually, models could potentially predict the outcome of
different treatment strategies and guide clinicians in the decision making.

In other parts of physical sciences and engineering, such as the
space industry or meteorology, the role of the computational
science has been crucial for its development. Without the increasing
computational power predicted by Moore's law
\cite{brock2006understanding} and better numerical methods, such as the
finite element method, scientist would not have been able to predict
complex physical phenomena such as a Tsunami, or being able to send
humans out in space.

Despite our ability to simulate complex physical systems such as
the weather, we still have a way to go before we are able to
simulate a human organ, such as the heart, with the same precision. The
multiscale nature of biophysical processes has challenged engineers for decades.
Even though we understand processes at each level, such as the flow of
ions in and out of the cell, or the electrical transduction through the
heart, the question of how to couple these levels together is still open
\cite{noble2002modeling}.

Another question is how to extract data
observed in medical images and constrain models, in order to make
models specific to the individual. These questions has to be addressed
before we can take the next step into personalized simulations for
\emph{in silico} medicine, and use computer simulations to guide
treatment, and predict outcome.

@@

The mechanics of the heart can be described via equations of nonlinear
elasticity which is traditionally formulated using a continuum
mechanics approach. This approach has been extensively used in other
parts of engineering, and numerical solutions to these partial
differential equations (PDE) are solved using the finite element
method (FEM) with the open source software package FEniCS \cite{logg2012automated}.
